Binning Strategies and Related Loss for Large Data

### 2015-04-08 Paper.Rnw
Commenting out old sections to clean up paper and speed up compiling

<<FreqLossDisplay,echo=F,include=F,eval=T,fig.width=15, fig.height=5, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="I am not sure that this figure is needed to demonstrate frequency binning. If so, do the colors between freq bins need to be made more distinguishable? i.e. in the quantile binning the bins starting at 1, 8 and 36.5 are hard to differentiate because the shade is determined by the breaks. ">>= 
binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)

@@

\subsection{Comparison of Loss in Binning Algorithms (old section - rework/remove soon)}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=TRUE>>=
# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
computeLoss <- function(binning, data){
  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
  lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
  datplot <- rbind(cbind(unique(lRandom$NewData), Algorithm="Random"), cbind(unique(lMean$NewData), Algorithm="Standard"))
  datplot$xbin <- round(datplot[,1]/binning[1],0)*binning[1]
  datplot$ybin <- round(datplot[,2]/binning[2],0)*binning[2]
  #datplot2 <- ddply(datplot, .(G, SO,Algorithm), summarize, count=length(G))
    qplot(xbin, ybin, geom="tile", data = datplot, fill = Freq , 
          ,main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
    ggsave(paste("./RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
  return(rbind(
    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$Loss), 
    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lMean$Loss)))
}

@@

\begin{figure}[hbtp]
\centering
<<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=

# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                      password="R R0cks", port=3306, dbname="baseball", 
#                      host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# d1 <- dbData(pitch, vars=c( "G", "SO"))
# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
# 
# library(multicore)
# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
# 
# lossesRdm <- cbind(bins, lossesRdm)
# lossesStd <- cbind(bins, lossesStd)
# 
# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)

losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
lossmelt$y <- as.factor(lossmelt$y)
levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
qplot(x=x, y=value,
      group=interaction(y, Algorithm),
      data=lossmelt, geom="line",
      colour=y, size=I(1),
      xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
        theme_bw() + theme(legend.position="bottom") + 
        guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
        facet_grid(variable ~ Algorithm, scales = "free")

# qplot(x=x, y=value,
#       group=interaction(y, Algorithm),
#       data=lossmelt, geom="line",
#       colour=y, size=I(1), linetype=Algorithm,
#       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
#         theme_bw() + theme(legend.position="bottom") + 
#         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
#         facet_grid(. ~ variable, scales = "free")

@
\caption{\label{fig:LossComparison1} Loss for Baseball Data under Standard and Random Binning for different bin widths in the number of strike-outs and games.}
\end{figure}

@@

} %end ktm color
%----------------------------------------------------------------------------

\subsection{Partition of Spatial Loss}\label{proof:partition}
We want to show that spatial loss can be written as a partition of the form 
\[L_s(x) = 1/S^2_{\emptyset}\sum_i S^2_i = 1/S^2_{\emptyset}\left[\sum_i L^N_i + \sum_i L^V_i\right]\]

For that let us consider the values $\{x_{i}: b(x_i) = x^\ast\}$ in a single bin with center $x^\ast$.
The numerical center of these points is given as  $1/n \sum x_i = \overline x$.
Then the total loss for each element $x_i$ in this bin is:

\begin{align*}
S_i^2 = \sum (x_i - x_i^\ast)^2 & = \sum (x_i - \overline x + \overline x - x_i^\ast)^2\\
& = \sum(x_i-\overline x)^2 + 2\sum (x_i-\overline x)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
& = \sum(x_i - \overline x)^2 + 2\sum(0)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
& = \sum(x_i - \overline x)^2 + \sum(\overline x-x_i^\ast)^2\\
& = L^N_i + L^V_i
\end{align*}
Adding over all elements in the bin and all bins gives the formula above.

\subsection{Constancy of Loss under Binary Splits}\label{proof:constantloss}
We want to show that under random binning the loss due to doubling of bin width is deterministic.

Let us consider the contribution $L_i$ of a single point $x_i$ to the loss when starting with the minimal bin width, i.e. each unique point is in a separate bin:
doubling the bin width leads to two  scenarios: 
a point is either at  the bin center or directly between two bins.
In the case that the point is at the bin center, its contribution to the total overall loss is zero,  $L_i = 0$, and the associated probability $p_i = 1$.

Alternately, when the point is exactly half-way between two bin centers, the point is attributed with probability $p_i = \frac{1}{2}$ to either bin, leading to a loss of $L_i = \frac{1}{2} c_i$ regardless of which bin is assigned. Hence, in this case, loss is also entirely independent of which bin is assigned.
\end{appendix}
\bibliographystyle{asa}
\bibliography{references}
\end{document}

>>>>

<<FreqLossDisplay,echo=F,include=T,eval=T,fig.width=15, fig.height=5, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="I am not sure that this figure is needed to demonstrate frequency binning. If so, do the colors between freq bins need to be made more distinguishable? i.e. in the quantile binning the bins starting at 1, 8 and 36.5 are hard to differentiate because the shade is determined by the breaks. ">>= 
binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)

@@

% \subsection{Comparison of Loss in Binning Algorithms (old section - rework/remove soon)}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=TRUE>>=
# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
computeLoss <- function(binning, data){
  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
  lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
  datplot <- rbind(cbind(unique(lRandom$NewData), Algorithm="Random"), cbind(unique(lMean$NewData), Algorithm="Standard"))
  datplot$xbin <- round(datplot[,1]/binning[1],0)*binning[1]
  datplot$ybin <- round(datplot[,2]/binning[2],0)*binning[2]
  #datplot2 <- ddply(datplot, .(G, SO,Algorithm), summarize, count=length(G))
    qplot(xbin, ybin, geom="tile", data = datplot, fill = Freq , 
          ,main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
    ggsave(paste("./RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
  return(rbind(
    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$Loss), 
    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lMean$Loss)))
}

@@

% \begin{figure}[hbtp]
% \centering
% <<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=
% 
% # connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
% #                      password="R R0cks", port=3306, dbname="baseball", 
% #                      host="headnode.stat.iastate.edu")
% # pitch <- new("dataDB", co=connect, table="Pitching")
% # d1 <- dbData(pitch, vars=c( "G", "SO"))
% # bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
% # 
% # library(multicore)
% # lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
% # lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
% # 
% # lossesRdm <- cbind(bins, lossesRdm)
% # lossesStd <- cbind(bins, lossesStd)
% # 
% # losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
% # write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
% 
% losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
% lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
% lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
% lossmelt$y <- as.factor(lossmelt$y)
% levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
% qplot(x=x, y=value,
%       group=interaction(y, Algorithm),
%       data=lossmelt, geom="line",
%       colour=y, size=I(1),
%       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
%         theme_bw() + theme(legend.position="bottom") + 
%         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
%         facet_grid(variable ~ Algorithm, scales = "free") 
% 
% # qplot(x=x, y=value,
% #       group=interaction(y, Algorithm),
% #       data=lossmelt, geom="line",
% #       colour=y, size=I(1), linetype=Algorithm,
% #       xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
% #         theme_bw() + theme(legend.position="bottom") + 
% #         guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
% #         facet_grid(. ~ variable, scales = "free") 
% 
% @
% \caption{\label{fig:LossComparison1} Loss for Baseball Data under Standard and Random Binning for different bin widths in the number of strike-outs and games.}
% \end{figure}

@@

%----------------------------------------------------------------------------
% 
% \subsection{Partition of Spatial Loss}\label{proof:partition}
% We want to show that spatial loss can be written as a partition of the form 
% \[L_s(x) = 1/S^2_{\emptyset}\sum_i S^2_i = 1/S^2_{\emptyset}\left[\sum_i L^N_i + \sum_i L^V_i\right]\] 
% 
% For that let us consider the values $\{x_{i}: b(x_i) = x^\ast\}$ in a single bin with center $x^\ast$.
% The numerical center of these points is given as  $1/n \sum x_i = \overline x$.
% Then the total loss for each element $x_i$ in this bin is:
% 
% \begin{align*}
% S_i^2 = \sum (x_i - x_i^\ast)^2 & = \sum (x_i - \overline x + \overline x - x_i^\ast)^2\\
% & = \sum(x_i-\overline x)^2 + 2\sum (x_i-\overline x)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
% & = \sum(x_i - \overline x)^2 + 2\sum(0)(\overline x-x_i^\ast) + \sum(\overline x-x_i^\ast)^2\\
% & = \sum(x_i - \overline x)^2 + \sum(\overline x-x_i^\ast)^2\\
% & = L^N_i + L^V_i
% \end{align*}
% Adding over all elements in the bin and all bins gives the formula above.
% 
% \subsection{Constancy of Loss under Binary Splits}\label{proof:constantloss}
% We want to show that under random binning the loss due to doubling of bin width is deterministic.
% 
% Let us consider the contribution $L_i$ of a single point $x_i$ to the loss when starting with the minimal bin width, i.e. each unique point is in a separate bin:
% doubling the bin width leads to two  scenarios: 
% a point is either at  the bin center or directly between two bins.
% In the case that the point is at the bin center, its contribution to the total overall loss is zero,  $L_i = 0$, and the associated probability $p_i = 1$. 
% 
% Alternately, when the point is exactly half-way between two bin centers, the point is attributed with probability $p_i = \frac{1}{2}$ to either bin, leading to a loss of $L_i = \frac{1}{2} c_i$ regardless of which bin is assigned. Hence, in this case, loss is also entirely independent of which bin is assigned.
\end{appendix}
\bibliographystyle{asa}
\bibliography{references}
\end{document}
